Estimating mismapping rates in seed-based algorithms for read mapping

### 2019-03-03 MEM.tex

@@

>>>>

The most popular heuristic for this problem is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm locally at those regions. The first step is a
filtration called ``seeding'' and the regions of high similarity are
called ``seeds''. This is for instance the general strategy of the popular
alignment algorithm BLAST~\cite{pmid2231712}.

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the (slow) alignment
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either $ii.$ or $iii.$, which are both erroneous.

The slowness of the alignment algorithm imposes a trade-off between speed
and accuracy. If the seeding step is set to produce many candidates, the
target is likely to be among them and it will be discovered. The downside
is that this will take long. Conversely, if the seeding step is set to
produce few candidates, the process will run faster but the target is less
likely to be in the candidate list so it is less likely to be discovered.

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others (see~\cite{pmid20460430,
sun2005designing,pmid11934743,xu2006optimizing,pmid17044164,
brejova2003vector,pmid18684737,pmid15359419} for examples of design,
see~\cite{pmid16533404, pmid20460430} for comparisons and discussions, and
see~\cite{navarro2001guided} for a more general review of filtration
methods in pattern matching). The upshot is that we can develop faster
mapping algorithms at no cost for accuracy if we find better seeding
strategies.

The seeding framework was developed for the best location problem, but it
also applies to the true location problem. Actually, the process is
exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
location. In outcome $i.$ called a ``true positive'', the DNA fragment is
mapped to the \emph{true} location. In outcome $ii.$ called a ``false
positive'', the DNA fragment is mapped to an incorrect location. Outcome
$iii.$ called a ``false negative'' is the same as before and means that
the DNA fragment is not mapped at all.

We will make two assumptions that will simplify the discussion. The first
is that the true location is always the best. We will further comment on
this and explain how to relax this assumption in section
section~\ref{sec:not_the_best}. Until then, we consider that the true
location problem is the same as the best location problem. The second
assumption is that the best location is always discovered if it is in the
candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
determined by the seeding process. For this reason, we will sometimes
refer to them as the outcomes of the seeding step, even though strictly
speaking they are the outcomes of the whole mapping pipeline\footnote{This
can be confusing because some authors refer to the off-target seeds as
false positives. When we say that the outcome of the seeding step is a
false positive, we mean that \emph{all} the seeds are off-target.}.

@@

In general, seeds do not have to be exact matches{}. For
instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

The rest of the document will focus on a seeding heuristic known as ``MEM
seeding'' (where MEM stands for Maximal Exact Match). This approach gives
good empirical results when mapping short reads (it is for instance the
strategy\footnote{To be exact, BWA-MEM uses SMEM seeds (Super Maximal
Exact Match seeds).} of the popular mapper BWA-MEM~\cite{li2013aligning})
but there is presently no theory to justify its performance. A deeper
understanding of MEM seeding would be useful to calibrate the
speed/accuracy trade-off of this heurisitic.
